---
title: "STS 122 - Lab 1"
author: "Roella Louise Pineda"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---
# Lab 1: Working with Data Frames
Due 1.8.21

## Scenario
You are the proprietor of a very small bookstore. You currently have the following books:

  - *The Waste Land* by T.S. Eliot (published 1922; 10 copies in stock; $8 each)  
  - *Lord of the Flies* by William Golding (published 1954; 16 copies in stock; $8 each)  
  - *Grapes of Wrath* by John Steinbeck (published 1939; 5 copies in stock; $10 each)  
  - *Pride and Prejudice* by Jane Austen (published 1813; 7 copies in stock; $15 each)  
  - *Frankenstein* by Mary Shelley (published 1818; 15 copies in stock; $5 each)  
  - *Little Women* by Louisa May Alcott (published 1868; 9 copies in stock; $20 each)  
  - *Jane Eyre* by Charlotte Bronte (published 1847; 12 copies in stock; $25 each)  
  - *The Things They Carried* by Tim O'Brien (published 1990; 2 copies in stock; $17 each)  
  - *Huckleberry Finn* by Mark Twain (published 1884; 21 copies in stock; $12 each)  
  - *Siddhartha* by Hermann Hesse (published 1922; 4 copies in stock; $22 each)

## Inventory
In the code block below, fill in the blanks to create a data frame with a row for each book and a column for each characteristic of the book (title, author, year, etc.). We will do this by creating a vector for each characteristic, then combining the vectors into a data frame. Print your result.
```{r}
title <- c("The Waste Land", "Lord of the Flies", "Grapes of Wrath", "Pride and Prejudice", "Frankenstein", "Little Women", "Jane Eyre", "The Things They Carried", "Huckleberry Finn", "Siddhartha" )
author <- c("T.S. Eliot", "William Golding", "John Steinbeck", "Jane Austen", "Mary Shelley", "Louisa May Alcott", "Charlotte Bronte", "Tim O'Brien", "Mark Twain", "Hermann Hesse" )
pyear <- c(1922, 1954, 1939, 1813, 1818, 1868, 1847, 1990, 1884, 1922)
nstock <- c(10, 16, 5, 7, 15, 9, 12, 2, 21, 4)
price <- c(8, 8, 10, 15, 5, 20, 25, 17, 12, 22)

bookstore <- data.frame(title, author, pyear, nstock, price)

bookstore
```

## Organization
For the remainder of the lab, we will be using functions from `dplyr`. Load the `dplyr` package in the code block below.
```{r}
library(dplyr)
```
Saving as a new data frame called `books_sorted`, fill in the blanks below to sort your inventory by year published (most recent first, so descending order) and, within year, by price (least expensive first, so ascending order). This can be done in either one step or two, but pay attention to the order. Print your result.
```{r}
books_sorted <- bookstore %>% arrange(-pyear, price)
books_sorted
```

## Sale
Your bookstore is having a sale!  

  - Books published before 1900 are 25% off  
  - Books with more than 10 copies in stock are 40% off  
  - Books published before 1900 with more than 10 copies in stock are 50% off

In the code block below, fill in the blanks to create a new data frame with the sale price for each book, in addition to the original price. You will need three ifelse statements, which can be either sequential or nested. Print your result.
```{r}
sale <- bookstore %>% mutate(saleprice =
ifelse(pyear < 1900 & nstock > 10, price * 0.50,
ifelse(pyear < 1900, price * 0.75,
ifelse(nstock > 10, price * 0.60, price))))
sale
```

## Sale price check
Add a column to the data frame with the sale prices that indicates the percentage discount for each book. Do this by calculating the difference between sale price and original price. Print your result.
```{r}
sale <- sale %>% mutate(percentdiscount = 
(price - saleprice)/price)
sale
```

## Sale list
Fill in the blanks below to create a new data frame that includes only books on sale and only author, title, and sale price. You will need to use the `filter()` and `select()` functions, and the order you use them in matters. Print your result.
```{r}
#This is my first answer. 
salebooks <- sale %>% 
  filter(percentdiscount >= 0.01) %>% 
  select(author, title, saleprice) 
salebooks

#Or is this the answer?
salebooks <- sale %>% 
  filter(percentdiscount >= 0.01) %>% 
  select(percentdiscount,author, title, saleprice) 
salebooks
```


## Turning in this assignment
Save and preview this file, then upload the `Lab1.nb.html` file to Canvas under Lab 1.